Constraining the Interiors of Asteroids Through Close Encounters

### 2022-08-16 paper/appendices.tex
Revised paper

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%% APPENDICES %%%%%%%%%%%%%%%%%%%%%

@@

In this appendix, we derive the equations of motion used to simulate the asteroid angular velocity during the encounter. In particular, we describe our coordinates (section \ref{sec:coordinates}) for an encountering asteroid's position and orientation, and we parametrize its density distribution via its density moments (section \ref{sec:moments}). Then we derive an arbitrary-order equation for tidal torque (section \ref{sec:tidal-torque}) and write the equations of motion for the system (section \ref{sec:eom}).

@@

The difference between the origins of the body-fixed and inertial frames is the position of the asteroid. We represent the relative orientations by $z-y-z$ Euler angles $\alpha$, $\beta$, and $\gamma$, such that a matrix $M$ rotating from the body-fixed to the inertial frame ($M\bm{r} = \bm{R}$) is given by
\begin{equation}
M = R_z(\alpha) R_y(\beta) R_z(\gamma).
\label{eqn:euler-angles}
\end{equation}
Here, $R_i(\theta)$ is a rotation around the unit vector $i$ by $\theta$ (figure \ref{fig:euler-angles}).

\begin{figure}
    \centering
    \begin{tikzpicture}
    \draw[-{Latex[length=3mm]}] (0, 0) -- (-4, 0) node[anchor=east] {$\unit x$};
    \draw[-{Latex[length=3mm]}] (0, 0) -- (2, -3) node[anchor=west] {$\unit y$};
    \draw[-{Latex[length=3mm]}] (0, 0) -- (0, 4) node[anchor=south] {$\unit z$};
    \draw[dashed, -{Latex[length=3mm]}] (0, 0) -- (3.7, -2) node[anchor=north] {};
    \draw[line width=0.5mm,-{Latex[length=3mm]}] (0, 0) -- (-0.5, -3) node[anchor=north] {$\unit X$};
    \draw[line width=0.5mm,-{Latex[length=3mm]}] (0, 0) -- (4, 1) node[anchor=south] {$\unit Y$};
    \draw[line width=0.5mm,-{Latex[length=3mm]}] (0, 0) -- (-1.6, 2.7) node[anchor=south] {$\unit Z$};
    \draw[->] (0.5, -0.75) arc (290:302:2.5);
    \draw (0.9, -0.9) node[anchor=center] {$\alpha$};
    \draw[->] (0, 1.2) arc (130:161:1.3);
    \draw (-0.3, 1.3) node[anchor=center] {$\beta$};
    \draw[->] (0.97, -0.52) arc (330:368:1.3);
    \draw (1.4, -0.25) node[anchor=center] {$\gamma$};
    %\draw[-{Latex[length=3mm]}] (0, 0) -- (0, 4) node[anchor=south] {$\unit Z$};
    %\draw[-{Latex[length=3mm]}] (0, 0) -- (4, 0) node[anchor=west] {$\unit Y$};
    %\draw[-{Latex[length=3mm]}] (0, 0) -- (-2, -2) node[anchor=east] {$\unit X$};
    \end{tikzpicture}
    \caption{$z-y-z$ Euler angles used in this work to express the orientation of the asteroid. Orientation is expressed as a rotation from the body-fixed axes (lowercase) to the inertial axes (bold lines and uppercase). The origins are co-located for demonstration purposes.}
    \label{fig:euler-angles}
\end{figure}

@@

Note that both $J_{\ell m}$ and $K_{\ell m}$ are unitless. We call them ``moments'' because the $R_{\ell m}(\bm r)$ contains an $r^\ell$ dependence so that $K_{\ell m}$ is the $\ell$th density moment of the asteroid.

@@

Finally, the requirement that $\rho_\mathcal{A}(\bm r) \geq 0$ everywhere restricts $K_{\ell m}$. In the case of $K_{2m}$, this fact and the constraint that $I_z$ is larger than $I_x$ or $I_y$ requires $K_{20}$ and $K_{22}$ to fall in the triangle
\begin{equation}
  -\frac{1}{4} \leq K_{20} \leq 0, \qquad |K_{22}| \leq -\frac{K_{20}}{2}.
  \label{eqn:parameter-bounds}
\end{equation}
An analytical constraint on $K_{3m}$ based on this property is more difficult to derive, but in practice, we also observe that $|K_{3m}| < 0.01$.

@@

The gravitational potential energy of the central body is, in its most general form,
\begin{equation}
V(\bm R') = -G\int_\mathcal{B} d^3 R \rho_\mathcal{B}(\bm R) \frac{1}{|\bm{R}-\bm{R'}|}.
\label{eqn:first-pe}
\end{equation}
where $\rho_\mathcal{B}$ is the density distribution of the central body and $\mathcal{B}$ indicates the central body's volume. All vectors here are written in the inertial frame. Given $|\bm{R}| < |\bm{R'}|$, Ref.~\cite{Gelderen1998TheSO} gives the identity
\begin{equation}
  \frac{1}{|\bm R - \bm R'|} = \sum_{\ell, m} R_{\ell m}(\bm R) S_{\ell m}^*(\bm R'),
  \label{eqn:ylm-expansion}
\end{equation}
where the sum is shorthand for $\sum_{\ell, m} = \sum_{\ell = 0}^\infty \sum_{m=-\ell}^\ell$.

@@

The equations of motion of the asteroid position $\bm D$ are given by Newton's law of gravitation
\begin{equation}
  \dot{\bm V} = -\frac{G \mu_\mathcal{B}}{D^3} \bm D \qquad \dot{\bm D} = \bm V
  \label{eqn:pos-eom}
\end{equation}
where $\bm V$ is the asteroid velocity in the inertial frame. Rather than derive equations of motion for the Euler angles (which suffer from gimbal lock), we instead represent the orientation of the asteroid with a quaternion $\quat q$ which can be converted into Euler angles to compute $\mathcal{D}(\alpha, \beta, \gamma)$. This quaternion evolves as 
\begin{equation}
  \dot{\quat q} = \frac{1}{2}\quat q\quat \omega.
  \label{eqn:quat-eom}
\end{equation}
for angular velocity $\bm \omega$ given in the body-fixed frame. The equations of motion of $\bm \omega$ in turn are given by
\begin{equation}
  \begin{split}
    I_x \dot \omega_x - \omega_y \omega_z (I_y - I_z) &= \tau_x\\
    I_y \dot \omega_y - \omega_z \omega_x (I_z - I_x) &= \tau_y\\
    I_z \dot \omega_z - \omega_x \omega_y (I_x - I_y) &= \tau_z.
  \end{split}
  \label{eqn:omega-eom}
\end{equation}
Equations \ref{eqn:tidal-torque}, \ref{eqn:moi}, and \ref{eqn:pos-eom} to \ref{eqn:omega-eom} form a set of non-linear, first-order coupled differential equations in which can be numerically integrated. They are expressed in terms of the physical parameters $I_\mathcal{A/B}$, $a_\mathcal{A/B}$, $J_{\ell m}$, and $K_{\ell m}$ which are constant if the asteroid is rigid and the central body does not rotate.

\section{Comparing orientation and angular velocity data}

@@

\subsection{Moment uncertainty comparison}
With the likelihood defined, we generate both angular velocity and orientation data for the asymmetric reference asteroid configuration and extract density moment PPDs for both data sets via the fit method defined in the main text. Due to the different uncertainty models used for the orientation and angular velocity data sets, this set-up does not allow direct comparison between the amount of moment uncertainty for both data sets. If one data set yields more precise moments than the other, one could not determine whether the effect is due to increased observational precision in the data set or the use of a data type that better constrains density moments. However, the relative uncertainty of moments can be compared between the two data sets.

To make this comparison, we compute moment uncertainties $\sigma(K_{\ell m})$ for both data sets. We then scale the moment uncertainties attained from the angular velocity data set so that the mean uncertainty of its parameters is equal to that of the orientation data set. This is equivalent to choosing observational uncertainties for the angular velocity data set which yield density moments to the same average precision as the observational data set, because figure \ref{fig:scan-observational} shows that density moment uncertainty $\sigma(K_{\ell m})$ is proportional to observational uncertainty. The resulting PPDs for the density moments of both data sets are displayed in figure \ref{fig:orientation-unc}, as well as their mean values relative to the true values.

\begin{figure}
  \centering
  \includegraphics[width=0.7\linewidth]{figs/orientation-unc}
  \caption{PPDs for each parameter as extracted from angular velocity (blue) and orientation (orange) data. First order parameters are shown in the top panel and second-order parameters in the bottom. Mean values are also shown as vertical lines. Both data produce similar constraints on parameters, except in the case of $\gamma_0$.}
  \label{fig:orientation-unc}
\end{figure}

@@

A smaller effect observed in figure \ref{fig:orientation-unc} is that the orientation data set yields similar uncertainties for all density moments of fixed $\ell$, whereas the angular velocity data set tends to yield larger uncertainties for small $|m|$. However, this has little effect on the density distributions extracted by the finite element model; the average density uncertainty $\sigma_\rho / \rho$ are essentially equivalent between the two data sets, as is the extracted distribution of density and density uncertainty.

@@

% \section{Dependence of the cadence cut-off on time scales}
% \label{app:cadence-tests}

% In section \ref{sec:scan-cadence}, we noted that moment uncertainty as a function of observation cadence $\Delta t$ increases suddenly near $\Delta t_\text{cut-off} \approx 30-40$ min. By dimensional analysis, $\Delta t_\text{cut-off}$ depends both on the asteroid period $P_\omega$ and the orbital time scales. In this appendix, we assess the relationship between $\Delta t_\text{cut-off}$ and these time scales by measuring moment uncertainty as a function of cadence at different values of these time scales. However, changing the time scales also affects the moment uncertainty directly, obscuring the cadence dependence of the moment uncertainty. Figure \ref{fig:scan-period} indicates that $P_\omega$ only slightly affects moment uncertainty as long as $P_\omega$ is not too small, so we neglect this obscuring effect in the case of $P_\omega$. To assess the dependence of $\Delta t$ on the orbital time scales, we fix the orbit shape and simulate the asteroid as following the orbit at different speeds. This process is unphysical, but it has the advantage of keeping all encounter parameters fixed except for the time the asteroid spends in the orbit, thereby isolating the orbital time scale. We parametrize this change $v_\text{orbit} / v_\text{physical}$, which is large for an artificially fast encounter, low for an artificially slow encounter, and 1 for a physical encounter.

% In figure \ref{fig:cad-contour}, we display contour plots of moment uncertainty $\sigma(K_{\ell m})$ of the fit parameters as a function of both cadence $\Delta t$ and $P_\omega$ and the relative orbit speed $v_\text{orbit} / v_\text{physical}$. Superimposed is the weak Both panels show the same sudden increase in posterior uncertainty we named $T_\text{cad}$, located around the region where $\sigma_\rho / \rho = 100\%$ and now visible as a function of frequency and relative orbit speed. In all cases, the value of $\gamma_0$ was set so that all data points achieve the same value of $\gamma$ at perigee.

% \begin{figure*}
%   \centering
%   \includegraphics[width=0.48\textwidth]{figs/cad-period.pdf}\hfill
%   \includegraphics[width=0.48\textwidth]{figs/cad-speed.pdf}
%   \caption{Contour plots showing posterior uncertainties as a function of cadence $\Delta t$ and dynamical time scales for the encounter: rotational period $P_\omega$ (\textit{left}) and the relative speed of the orbit (\textit{right}; see text for a definition). The reference values of $P_\omega=9$ hr and $t_\text{spin}/t_\text{orbit}=1$ are shown as dotted lines. The solid (dotted) red line represents the $\sigma_\rho / \rho = 100\%$ (20\%) threshold. Cadence cut-off depends strongly on both $P_\text{omega}$ and $t_\text{spin}/t_\text{orbit}=1$.}
%   \label{fig:cad-contour}
% \end{figure*}

% Figure \ref{fig:cad-contour} demonstrates that large rotational period produces high $T_\text{cad}$. The dependence on $P_\omega$ agrees with the fact that large rotational periods for fixed cadence lead to better posterior uncertainty, discussed in section \ref{sec:scan-period}. The figure also demonstrates that for large $P_\omega$, $T_\text{cad}$ depends less strongly on $P_\text{omega}$ and may even reverse its dependence such that increasing $P_\text{omega}$ decreases $T_\text{omega}$. In all cases except $\Re K_{33}$, at least constant-$\sigma(K_{\ell m})$ contour is seen to curve back such that $\Delta t$ decreases as a function of $P_\text{omega}$ for large $P_\text{omega}$. In these regions the cadence cut-off also dulls, as shown by the spreading of the constant-$\sigma(K_{\ell m})$ contours in this region.

%  The relative orbit speed $t_\text{spin} / t_\text{orbit}$ was defined by un-physically increasing or decreasing the time at which the asteroid moved through the orbit determined by the equations of motion (but leaving the orbit shape unchanged). The equations of motion affecting the orientation and spin of the asteroid however were unaffected. $t_\text{spin} / t_\text{orbit} > 1$ corresponds to a faster orbit, and $t_\text{spin} / t_\text{orbit} < 1$ corresponds to a slower orbit. With this unphysical process, we isolate the effect of the amount of time spent near perigee on posterior uncertainty, without inheriting additional affects that would have been caused by the orbit changing shape.

% The right panel of figure \ref{fig:cad-contour} shows a stronger and more monotonic dependence of $T_\text{cad}$ on $t_\text{spin} / t_\text{orbit}$. It appears that even slightly slower orbits sharply increases $T_\text{cad}$. This effect is both due to the asteroid spending greater time in the high-torque, near-perigee region, and the larger data set that can be collected for slow orbits. However, if the orbit speed is changed by adjusting its parameters ($v_\infty$, $r_p$) or the central body mass $\mu_\mathcal{B}$, then the orbit shape will also change. This induces other effects studied in the main text and will complicate the trend observed here. Unlike the $P_\omega$ case, the cadence cut-off does not visibly broaden as a function of $t_\text{spin} / t_\text{orbit}$. It appears that the increased orbit speed merely shifts $T_\text{cad}$ rather than changing its sharpness.

>>>>

@@

In this appendix, we derive the equations of motion used to simulate the asteroid encounter. In particular, we describe our coordinates (section \ref{sec:coordinates}) for an encountering asteroid's position and orientation, and we parametrize its density distribution via its density moments (section \ref{sec:moments}). Then we derive an arbitrary-order equation for tidal torque (section \ref{sec:tidal-torque}) and write the equations of motion for the system (section \ref{sec:eom}).

@@

The difference between the origins of the body-fixed and inertial frames is the position of the asteroid. We represent the relative orientations by $z-y-z$ Euler angles $\alpha$, $\beta$, and $\gamma$, such that a matrix $M$ rotating from the body-fixed to the inertial frame ($M\bm{r} = \bm{R}$) is given by
\begin{equation}
M = R_z(\alpha) R_y(\beta) R_z(\gamma).
\label{eqn:euler-angles}
\end{equation}
Here, $R_i(\theta)$ is a rotation around the unit vector $i$ by angle $\theta$ (figure \ref{fig:euler-angles}).

\begin{figure}
    \centering
    \begin{tikzpicture}
    \draw[-{Latex[length=3mm]}] (0, 0) -- (-4, 0) node[anchor=east] {$\unit x$};
    \draw[-{Latex[length=3mm]}] (0, 0) -- (2, -3) node[anchor=west] {$\unit y$};
    \draw[-{Latex[length=3mm]}] (0, 0) -- (0, 4) node[anchor=south] {$\unit z$};
    \draw[dashed, -{Latex[length=3mm]}] (0, 0) -- (3.7, -2) node[anchor=north] {};
    \draw[line width=0.5mm,-{Latex[length=3mm]}] (0, 0) -- (-0.5, -3) node[anchor=north] {$\unit X$};
    \draw[line width=0.5mm,-{Latex[length=3mm]}] (0, 0) -- (4, 1) node[anchor=south] {$\unit Y$};
    \draw[line width=0.5mm,-{Latex[length=3mm]}] (0, 0) -- (-1.6, 2.7) node[anchor=south] {$\unit Z$};
    \draw[->] (0.5, -0.75) arc (290:302:2.5);
    \draw (0.9, -0.9) node[anchor=center] {$\alpha$};
    \draw[->] (0, 1.2) arc (130:161:1.3);
    \draw (-0.3, 1.3) node[anchor=center] {$\beta$};
    \draw[->] (0.97, -0.52) arc (330:368:1.3);
    \draw (1.4, -0.25) node[anchor=center] {$\gamma$};
    \end{tikzpicture}
    \caption{$z-y-z$ Euler angles used in this work to express the orientation of the asteroid. Orientation is expressed as a rotation from the body-fixed axes (lowercase) to the inertial axes (bold lines and uppercase). The origins are co-located for demonstration purposes.}
    \label{fig:euler-angles}
\end{figure}

@@

Note that both $J_{\ell m}$ and $K_{\ell m}$ are unitless. We call them ``moments'' because $R_{\ell m}(\bm r)$ contains an $r^\ell$ dependence so that $K_{\ell m}$ is the $\ell$th density moment of the asteroid.

@@

Finally, the requirement that $\rho_\mathcal{A}(\bm r) \geq 0$ everywhere restricts $K_{\ell m}$. In the case of $K_{2m}$, this fact and the constraint that $I_z$ is larger than $I_x$ or $I_y$ requires $K_{20}$ and $K_{22}$ to fall in the triangle
\begin{equation}
  -\frac{1}{4} \leq K_{20} \leq 0, \qquad |K_{22}| \leq \frac{|K_{20}|}{2}.
  \label{eqn:parameter-bounds}
\end{equation}
An analytical constraint on $K_{3m}$ based on this property is more difficult to derive, but in practice, we observe that $|K_{3m}| < 0.01$.

@@

The gravitational potential energy of the central body is, in its most general form,
\begin{equation}
V(\bm R') = -G\int_\mathcal{B} d^3 R \frac{\rho_\mathcal{B}(\bm R)}{|\bm{R}-\bm{R'}|}.
\label{eqn:first-pe}
\end{equation}
where $\rho_\mathcal{B}$ is the density distribution of the central body and $\mathcal{B}$ indicates the central body's volume. All vectors here are written in the inertial frame. Given $|\bm{R}| < |\bm{R'}|$, Ref.~\cite{Gelderen1998TheSO} gives the identity
\begin{equation}
  \frac{1}{|\bm R - \bm R'|} = \sum_{\ell, m} R_{\ell m}(\bm R) S_{\ell m}^*(\bm R'),
  \label{eqn:ylm-expansion}
\end{equation}
where the sum is shorthand for $\sum_{\ell, m} = \sum_{\ell = 0}^\infty \sum_{m=-\ell}^\ell$.

@@

The equations of motion of the asteroid position $\bm D$ are given by Newton's law of gravitation
\begin{equation}
  \dot{\bm V} = -\frac{G \mu_\mathcal{B}}{D^3} \bm D \qquad \dot{\bm D} = \bm V
  \label{eqn:pos-eom}
\end{equation}
where $\bm V$ is the asteroid velocity in the inertial frame. Rather than derive equations of motion for the Euler angles (which suffer from gimbal lock), we instead represent the orientation of the asteroid with a quaternion $\quat q$ which can be converted into Euler angles to compute $\mathcal{D}(\alpha, \beta, \gamma)$. This quaternion evolves as 
\begin{equation}
  \dot{\quat q} = \frac{1}{2}\quat q\quat \omega.
  \label{eqn:quat-eom}
\end{equation}
for angular velocity $\bm \omega$ given in the body-fixed frame. The equations of motion of $\bm \omega$ in turn are given by
\begin{equation}
  \begin{split}
    I_x \dot \omega_x - \omega_y \omega_z (I_y - I_z) &= \tau_x\\
    I_y \dot \omega_y - \omega_z \omega_x (I_z - I_x) &= \tau_y\\
    I_z \dot \omega_z - \omega_x \omega_y (I_x - I_y) &= \tau_z.
  \end{split}
  \label{eqn:omega-eom}
\end{equation}
Equations \ref{eqn:tidal-torque}, \ref{eqn:moi}, \ref{eqn:pos-eom}, and \ref{eqn:omega-eom} form a set of non-linear, first-order coupled differential equations in which can be numerically integrated. They are expressed in terms of the physical parameters $I_\mathcal{A/B}$, $a_\mathcal{A/B}$, $J_{\ell m}$, and $K_{\ell m}$ which are constant if the asteroid is rigid and the central body does not rotate.

\section{Comparing orientation and angular velocity data}
\label{app:orientation}

@@

\subsection{Moment uncertainty comparison}
With the likelihood defined, we generate both angular velocity and orientation data for the asymmetric reference asteroid configuration and extract density moment PPDs for both data sets via the fit method defined in the main text. Due to the different uncertainty models used for the orientation and angular velocity data sets, this set-up does not allow direct comparison between the amount of moment uncertainty for both data sets. (If one data set yields more precise moments than the other, one could not determine whether the effect is due to increased observational precision in the data set or the use of a data type that better constrains density moments.) However, the relative uncertainty of moments can be compared.

To make this comparison, we compute moment uncertainties $\sigma(K_{\ell m})$ for both data sets. We then scale the moment uncertainties attained from the angular velocity data set so that the mean uncertainty of its parameters is equal to that of the orientation data set. This is equivalent to choosing observational uncertainties for the angular velocity data set which yield density moments to the same average precision as the observational data set. The resulting PPDs for the density moments of both data sets are displayed in figure \ref{fig:orientation-unc} relative to the true values.

\begin{figure}
  \centering
  \includegraphics[width=0.7\linewidth]{figs/orientation-unc}
  \caption{PPDs for each parameter as extracted from angular velocity (blue) and orientation (orange) data. First order parameters are shown in the top panel and second-order parameters in the bottom. Mean values are also shown as vertical lines. Both data sets produce similar constraints on parameters, except in the case of $\gamma_0$.}
  \label{fig:orientation-unc}
\end{figure}

@@

A smaller effect observed in figure \ref{fig:orientation-unc} is that the orientation data set yields similar uncertainties for all density moments of fixed $\ell$, whereas the angular velocity data set tends to yield larger uncertainties for small $|m|$. However, this has little effect on the density distributions extracted by the finite element model (not shown); the average density uncertainty $\sigma_\rho / \rho$ are essentially equivalent between the two data sets, as is the extracted distribution of density and density uncertainty.

@@
